Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-06-24 paper/document.tex
Update document text and add balance.

@@

A variety of visualizations tools have been created in order to help with understanding the changes that get made to software over time. Wu \textit{et al.}~\cite{wu2004a} explored punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.}~\cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies~\cite{6980224} or files~\cite{6650525, 7332421, voinea2006} over time.

@@

Some tools examine the evolution of projects to learn other information about them. Alcocer \textit{et al.}~\cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ogawa and Ma created code\_swarm~\cite{codeswarm}, a generative art visualization which presents the progressive evolution of file changes made by authors in a source control repository over time. Gource~\cite{gource} is a similar visualization, showing an animated force-directed tree of the file hierarchy as authors make changes to files in the source code repository over time. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files~\cite{6980223}. This allowed testing practices to be more closely examined. Anslow \textit{et al.}~\cite{sourcevis} created SourceVis, a suite of software visualizations including visualizations for semantically-aware project evolution. SourceVis displays the evolution of Java source code metrics such as number of packages, classes, methods, and fields.

None of the aforementioned tools perform comparisons of ASTs or specifically examine the changes to type information for their view of project evolution.

@@

Developers using the massively popular GitHub website for hosting their Git repositories instantly have access to basic visualizations, known as GitHub Graphs (see Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. GitHub Graphs tracks contributions based only on the number of commits and number of lines changed. This has the advantage of not relying on the code which was committed. It does not matter what language is used or if errors exist in the commit---GitHub will simply output the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear idea of what is being changed, only that changes have occurred. This data may be lacking, or even misleading which contributes to the need for better tools.

@@

Given a Git repository, we want to extract per-author type changes over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}). This extracted data can then be used in a variety of visualizations as discussed in Section~\ref{sec:viz}. The reason for using ASTs is that they capture the semantics of the code far better than simple line counts.

An \emph{abstract syntax tree} (AST) is a data structure derived from source code that breaks its syntactic constructs into a tree. Each node in the tree represents a syntactically valid chunk of code. This captures the structure of the code in an abstract and tractable manner. When a programmer makes a non-trivial change in code in a file, that difference will be reflected in the AST\@. Non-trivial changes are those that affect how a compiler will interpret the source code, and may have effects on the executable code. Trivial changes are those that have no effect on how a compiler will interpret the source text, such as changing insignificant whitespace or comments in source code. Thus, ASTs of consecutive revisions in a code repository can be compared to see what \emph{structures} a programmer is modifying.

@@

To determine what has changed between two subsequent commits, we computed their \emph{AST difference}. Given two ASTs, an AST difference (or simply \emph{AST diff}) states which nodes must be added and which nodes must be removed to transform one tree to the next (step two in Figure~\ref{fig:context}). This allows us to precisely track changes between two consecutive revisions of the same file. For AST diffing, we used GumTree~\cite{falleri:hal-01054552}, an algorithm made to compute the difference between two ASTs generated by Spoon. For new files and deleted files, it is unnecessary to calculate an AST difference; we simply run Spoon on the files and treated everything as an addition or deletion respectively.

@@

This states that two declarations of type \texttt{java.net.URI} were added, and two declarations of type \texttt{java.lang.String} were deleted. Note the fully qualified type names. The source text may not literally have the characters \texttt{java.net.URI}, but Spoon was able to infer from the import statements in the Java source text that references to ``URI'' resolve to the fully-qualified type name \texttt{java.net.URI}. Thus, TypeV is able to keep track of the actual class being used and is not confused when two classes in separate packages have the same name.

@@

We deliberately chose to ignore merge commits. A merge commit has two cases to consider: The first case is when the parents of the commit can be automatically merged. In this case all the relevant changes are already recorded in the parent of the commit so we can safely ignore the merge commit~\cite{git:merge}. The second case occurs when the parents cannot be merged automatically: this is a \emph{merge conflict} which requires manual resolution on the part of a developer. In this case, the developer will try to resolve the issue by modifying code. If we decide to include the merge commit we run into several edge case issues such as how to compare the merge commit with its multiple parent commits and how to account for the possible modifications made to resolve the merge conflict. We could consider comparing the merge commit with only one parent commit at a time; however, we will end up counting all the changes already recorded in the other parent branch. Additionally, all changes made during the resolution of a conflict will be attributed to the author making the merge. This may exclude the author that actually wrote the code in the first place. The other option is to simply ignore the merge commit such that we do not count the changes done during a merge commit. We made the assumptions that merge conflicts occur infrequently and when they do occur, the changes made to resolve a merge conflict are usually minor. Therefore, we chose to ignore merge commits. Note that fast-forwards are not \emph{true merges}, hence they do not create merge commits~\cite{git:merge} so none of the previous cases involving merges apply.

@@

Although rich in detail, the AST diff data is difficult to analyze if displayed in its raw form; hence, we created an HTML5 web visualization application which is intended for developers, project managers, and researchers to use (Figure~\ref{fig:main-view}). The source code for TypeV---both the AST differencing component and the web application---is available online on GitHub.\footnote{\url{https://github.com/mdfeist/TypeV}} A screencast demonstrating its use is also available online.\footnote{\url{https://archive.org/details/typev-demo}}

@@

The stacked bar graphs succinctly displays the relative amounts of additions and deletions per each time period. This creates a dense view of the project's evolution. We chose to make the height of the bars relative to only the type itself to make it easy for a user to scan a single type and see when it was modified. This makes tasks such as finding a refactoring (see Section~\ref{sec:antlr4}) very clear to see simply by browsing the line. We stacked the additions on top of the deletions such that users can immediately compare the relative frequency of changes of a single type throughout the chosen time span.

@@

By adding interaction, the user is able to explore the timeline to find the specific information which they are interested in and drill-down at their leisure. Although our tool extracts a great deal of information, by using interaction, users can gradual reveal more information as they need it, rather than be bombarded with all of the information at once.

\section{Evaluation}

@@

\begin{table}[t]
\renewcommand{\arraystretch}{1.3}
\centering
\caption{Summary of projects used in the evaluation.}
\label{tab:project-summary}
\begin{tabular}{c|ccccc}
\hline
\bfseries Project & \bfseries Commits & \bfseries Authors & \bfseries Files &   \bfseries Types &  \bfseries  Changes\\
\hline
ANTLR4 & 3930 & 27 & 1793 & 2635 & 58425\\
BookKeeper & 660 & 12 & 758 & 1872 & 31817\\
Curator & 1524 & 43 & 1211 & 1269 & 27957\\
Tika & 860 & 10 & 604 & 757 & 20764\\
\hline
\multicolumn{6}{p{3.25in}}{\footnotesize Totals for the four projects we analyzed. ``Changes'' refers to the total number of invocation or declaration changes observed from the AST diffs. Note that the number given for authors may be larger than the actual number of collaborators on the project, due to some collaborators using multiple email address, thus appearing as more than one author.}
\end{tabular}
\end{table}

@@

Is it worth the effort of computing commit-by-commit AST differences if we can just simply count changes to files? We define \emph{type coverage} as the number of types that have been touched out of the total number of types during either a certain time period or the entire project. \emph{File coverage} is the same measurement, but using the number of source files touched instead. File coverage can be trivially determined from an author's commit history. Type coverage, however, is uniquely provided by AST differencing, as we have described.

@@

To determine if type coverage yields different information than file coverage, we analyzed four open source Java projects: ANTLR4, Apache BookKeeper, Apache Curator and Apache Tika. Three of the four projects were collected from the Apache foundation, as they host a large number of well-curated open source Java projects. A summary of the projects studied is given in Table~\ref{tab:project-summary}. We counted the file coverage and type coverage on a commit-by-commit basis for each author. That is, each datum represents a commit with three dimensions: file coverage, type coverage, and its commit date. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the statistics measured.

We displayed the density distributions for two different projects: ANTLR4 (Figure~\ref{fig:antlr4-density}) and Apache Tika (Figure~\ref{fig:tika-density}). These plots show the distribution of the amount of commits that have X\% of type and file coverage. For example, Figure~\ref{fig:tika-density} shows that there are many commits with 50\% type coverage and very few commits with 50\% file coverage. This shows that even though the type and file coverage plots may have a similar shape the commits are showing different levels of information.

@@

We conclude from this that type coverage yields related, yet different information than file coverage alone.

\section{Illustrative Case Studies}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\columnwidth]{ParserATNSimulator}
\caption{Snapshot of TypeV showing changes to \texttt{ParserATNSimulator} in ANTLR4. Date ranges from approximately the beginning of 2011 to the end of April, 2012.}
\label{fig:parser1}
\end{figure}

\begin{figure}[!ht]
\centering
\includegraphics[width=\columnwidth]{Antlr4-GH-Lines-ParserATNSimulator}
\caption{GitHub's code frequency graph (based on line difference) showing changes to ANTLR4. The time spans approximately the same length as Figure~\ref{fig:parser1}.}
\label{fig:parser2}
\end{figure}

@@

For example, in December of 2011 we can see that the \texttt{ParserATNSimulator} class has a large amount of additions and deletions in roughly equal proportion (Figure~\ref{fig:parser1}). This suggests that \texttt{ParserATNSimulator} is not being removed from the project but instead there is some refactoring being done. If we take a closer look at the commit messages we can see that they mention refactoring and reorganization of code. Zooming in on that date,  shows that during that month they were finishing work on a new version of \texttt{ParserATNSimulator}.

If we tried to do the same analysis with the GitHub tools it would be significantly more difficult to reach this conclusion (Figure~\ref{fig:parser2}). First, if we only look at line changes this change to \texttt{ParserATNSimulator} is overwhelmed by other changes and hardly shows up on the GitHub tools. If we did notice an increase in activity we would not know what types were being worked on. Finally, we could not quickly look at the commit messages for that specific time and type if we used the GitHub tools.

>>>>

% Copyright (C) 2016 Michael D. Feist, Eddie Antonio Santos, Ian Watts, Abram Hindle
% 
% All TeX sources are licensed under Creative Commons Attribution-NoDerivatives 4.0 International (CC BY-ND 4.0) license.
% See terms at <http://creativecommons.org/licenses/by-nd/4.0/>

@@

Several visualizations tools have been created to understand how a software system changes over time. Wu \textit{et al.}~\cite{wu2004a} explored punctuated changes of software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.}~\cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies~\cite{6980224} or files~\cite{6650525, 7332421, voinea2006} over time. Alcocer \textit{et al.}~\cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ogawa and Ma created code\_swarm~\cite{codeswarm}, a generative art visualization which presents the progressive evolution of file changes made by authors in a source control repository over time. Gource~\cite{gource} is a similar visualization, showing an animated force-directed tree of the file hierarchy as authors make changes to files in the source code repository over time. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files~\cite{6980223}. This allowed testing practices to be more closely examined. Anslow \textit{et al.}~\cite{sourcevis} created SourceVis, a suite of software visualizations including visualizations for semantically-aware project evolution. SourceVis displays the evolution of Java source code metrics such as number of packages, classes, methods, and fields.

@@

@@

Developers using the massively popular GitHub website for hosting their Git repositories instantly have access to basic visualizations, known as GitHub Graphs (Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. GitHub Graphs tracks contributions based only on the number of commits and number of lines changed. This has the advantage of not relying on the code which was committed. It does not matter what language is used or if errors exist in the commit---GitHub will simply output the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear idea of what is being changed, only that changes have occurred. This data may be lacking, or even misleading which contributes to the need for better tools.

@@

Given a Git repository, we want to extract per-author use of types over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}). This extracted data can then be used in a variety of visualizations as discussed in Section~\ref{sec:viz}. The reason for using ASTs is that they capture the semantics of the code far better than simple line counts.

An \emph{abstract syntax tree} (AST) is a data structure derived from source code that breaks its syntactic constructs into a tree. Each node in the tree represents a syntactically valid chunk of code. This captures the structure of the code in an abstract and tractable manner. When a programmer makes a non-trivial change to a source code file, that difference will be reflected in the AST\@. Non-trivial changes are those that affect how a compiler will interpret the source code, and may have effects on the executable code. Trivial changes are those that have no effect on how a compiler will interpret the source text, such as changing insignificant whitespace or comments in source code. Thus, ASTs of consecutive revisions in a code repository can be compared to see what \emph{structures} a programmer is modifying.

@@

To determine what has changed between two subsequent commits, we computed their \emph{AST difference}. Given two ASTs, an AST difference (or simply as \emph{AST diff}) states which nodes must be added and which nodes must be removed to transform one tree to the next (step two in Figure~\ref{fig:context}). This allows us to precisely track changes between two consecutive revisions of the same file. For AST diffing, we used GumTree~\cite{falleri:hal-01054552}, an algorithm made to compute the difference between two ASTs generated by Spoon. For new files and deleted files, it is unnecessary to calculate an AST difference; we simply run Spoon on the files and treated everything as an addition or deletion respectively.

@@

This states that two declarations of type \texttt{java.net.URI} were added, and two declarations of type \texttt{java.lang.String} were deleted. Note that TypeV is able to find the fully qualified type names. Thus, the source text may not literally have the characters \texttt{java.net.URI}, but Spoon was able to infer from the import statements in the Java source text that references to ``URI'' resolve to the fully-qualified type name \texttt{java.net.URI}.

@@

We deliberately chose to ignore merge commits. A merge commit has two cases to consider: The first case is when the parents of the commit can be automatically merged. In this case all the relevant changes are already recorded in the parent of the commit so we can safely ignore the merge commit~\cite{git:merge}. The second case occurs when the parents cannot be merged automatically: this is a \emph{merge conflict} which requires manual resolution on the part of a developer. In this case, the developer will try to resolve the issue by modifying code. If we decide to include the merge commit we run into several edge case issues such as how to compare the merge commit with its multiple parent commits and how to account for the possible modifications made to resolve the merge conflict. We could consider comparing the merge commit with only one parent commit at a time; however, we will end up counting all the changes already recorded in the other parent branch. Additionally, all changes made during the resolution of a conflict will be attributed to the author making the merge. This may exclude the author that actually wrote the code in the first place. The other option is to simply ignore the merge commit such that we do not count the changes done during a merge commit. We made the assumptions that merge conflicts occur infrequently and when they do occur, the changes made to resolve a merge conflict are usually minor. Therefore, we chose to ignore merge commits.

@@

Although rich in detail, the AST diffs are difficult to analyze if displayed in their raw form; hence, we created an HTML5 web visualization application which is intended for developers, project managers, and researchers to use (Figure~\ref{fig:main-view}). The source code for TypeV---both the AST differencing component and the web application---is available online on GitHub.\footnote{\url{https://github.com/mdfeist/TypeV}} A screencast demonstrating its use is also available online.\footnote{\url{https://archive.org/details/typev-demo}}

@@

The stacked bar graphs succinctly displays the relative amounts of additions and deletions per each time period. This creates a dense view of the project's evolution. We chose to make the height of the bars relative to only the type itself to make it easy for a user to scan a single type and see when it was modified. This makes tasks such as finding a refactoring (Section~\ref{sec:antlr4}) very clear to see simply by browsing the line. We stacked the additions on top of the deletions such that users can immediately compare the relative frequency of changes of a single type throughout the chosen time span.

@@

By adding interaction, the user is able to explore the timeline to find the specific information that they are interested in and drill-down at their leisure. Although our tool extracts a great deal of information, users can gradual reveal more information as they need it, rather than be bombarded with all of the information at once.

\section{Evaluation}
\label{sec:eval}

@@

\begin{table}[t]
\renewcommand{\arraystretch}{1.3}
\centering
\caption{Summary of projects used in the evaluation.}
\label{tab:project-summary}
\begin{tabular}{c|ccccc}
\hline
\bfseries Project & \bfseries Commits & \bfseries Authors & \bfseries Files &   \bfseries Types &  \bfseries  Changes\\
\hline
ANTLR4 & 3930 & 19 & 1792 & 2635 & 58425\\
BookKeeper & 660 & 10 & 758 & 1872 & 31817\\
Curator & 1524 & 35 & 1211 & 1269 & 27957\\
Tika & 860 & 10 & 604 & 757 & 20764\\
\hline
\multicolumn{6}{p{3.25in}}{\footnotesize Totals for the four projects we analyzed. ``Changes'' refers to the total number of invocation or declaration changes observed from the AST diffs.}
\end{tabular}
\end{table}

@@

\textbf{RQ2}: Is it worth the effort of computing commit-by-commit AST differences if we can simply count changes to files? We will answer this question by comparing type coverage and file coverage. We define \emph{type coverage} as the number of types that have been touched out of the total number of types during either a certain time period or the entire project. \emph{File coverage} is the same measurement, but using the number of source files touched instead. File coverage can be trivially determined from an author's commit history. Type coverage, however, is uniquely provided by AST differencing, as we have described.

@@

To determine if type coverage yields different information than file coverage, we analyzed four open source Java projects: ANTLR4, Apache BookKeeper, Apache Curator and Apache Tika (Figure~\ref{tab:project-summary}). Three of the four projects were collected from the Apache foundation, as they host a large number of well-curated open source Java projects. We counted the file coverage and type coverage on a commit-by-commit basis for each author. That is, each datum represents a commit with three dimensions: file coverage, type coverage, and its commit date. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the statistics measured.

We displayed the density distributions for two different projects: ANTLR4 (Figure~\ref{fig:antlr4-density}) and Apache Tika (Figure~\ref{fig:tika-density}). These plots show the distribution of the amount of commits that have X\% of type and file coverage. For example, Figure~\ref{fig:tika-density} shows that there are many commits with 50\% type coverage and while there are relatively few commits with 50\% file coverage. This shows that even though the type and file coverage plots may have a similar shape, the commits are showing different levels of information.

@@

We conclude from these findings that type coverage yields related, yet different information than file coverage alone.

\section{Illustrative Case Studies}
\label{sec:cases}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\columnwidth]{ParserATNSimulator-Antlr4-GH-Lines}
\caption{Screenshots of TypeV (above) and GitHub's code frequency graph based on line difference (below). TypeV is showing changes to \texttt{ParserATNSimulator} in ANTLR4 while GitHub's graph is showing all line changes to ANTLR4. The date ranges from approximately April 2011 to April 2012. Note that both the TypeV and GitHub graphs are approximately lined up along the x-axis (time) and that we can see the loss of detail in the GitHub graphs.}
\label{fig:parser}
\end{figure}

@@

For example, in December of 2011, in Figure~\ref{fig:parser} we can see that the \texttt{ParserATNSimulator} class has a large amount of additions and deletions in roughly equal proportion. This suggests that \texttt{ParserATNSimulator} is not being removed from the project but instead there is some refactoring being done. If we take a closer look at the commit messages we can see that they mention refactoring and reorganization of code. Zooming in on that date,  shows that during that month they were finishing work on a new version of \texttt{ParserATNSimulator}.

If we tried to do the same analysis with the GitHub tools it would be significantly more difficult to reach this conclusion (Figure~\ref{fig:parser}). First, if we only look at line changes this change to \texttt{ParserATNSimulator} is overwhelmed by other changes and hardly shows up on the GitHub tools. If we did notice an increase in activity we would not know what types were being worked on. Finally, we could not quickly look at the commit messages for that specific time and type if we used the GitHub tools.
